A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-31 paper.tex
Create Big and Little ARG appendix
Move related content in there, without actually doing any new writing

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of species \citep{rannala2003genetics}, non (or
rarely) recombining sections of DNA \citep{cann1987mitochondrial,underhill2001annalsofhumangenetics} and viruses \citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line}
% Do we mean "despite the fact that analysis of genetic data relies on methods
% which ultimately derive from the inheritance paths defined by genetic trees
% and the mutations upon them, analysis so far has focussed on summaries ...
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

@@

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\item Throughout, we consider ARGs defined for a single chromosome comprising
of a fixed number of discrete sites. Our argument applies equally well to ARGs
defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
koskela2019robust} and to idealised, continuous chromosomes, but addressing
all these cases simultaneously would result in needlessly bulky notation.
\end{enumerate}

@@

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph
(with particular properties; see section XXX for details),
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details).

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now broken,
since such parsimonious ARGs are extremely unlikely under the
% JK: I assume this is true; any citations to back it up I wonder?
coalescent with recombination.
\cite{minichiello2006mapping} made this separation of the ARG as a
stochastic process vs the ARG as a data structure explicit, and
for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as examining more general phylogenetic
networks~\citep{huson2010phylogenetic}.

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
However, in parallel with this tendancy to decouple the term
Ancestral Recombination graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model an ensemble of correlated
potential ancestral trees and has been of considerable
and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
A vital distinction between the ARG (process) and ASG, however, is that there
is that there is a hard distinction between the stochastic process
(the graph) and an observable realisation (a single tree) because
the ASG encodes \emph{potential} ancestral trees.
Similarly, the Ancestral Gene Transfer Graph
models horizontal gene transfer~\citep{baumdicker2014infinitely}
by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form was it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely derived from the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces spurious precision about the location and timing
of recombination events.

% Although Hudson's algorithm was accompanied by a graph
% schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
% we will use ARG to refer to a graph structure, and refer to the
% process as the Griffiths process or the coalescent with recombination
% \citep{minichiello2006mapping}.

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
% to go with this
% In the Griffiths formulation, each edge in the graph corresponds to an extant
% lineage and nodes are events in the process (the initial $n$ leaf nodes are
% ``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
% are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
% into a common ancestor lineage. Recombination events happen at a rate of
% $k \rho$. Here, we choose a lineage (edge) uniformly, and a
% breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
% node, record the breakpoint and start two new edges from this node. The process
% then continues until there is only one lineage left (the Grand Most Recent
% Common Ancestor, GMRCA), which is guaranteed to
% happen in finite time because of the linear vs quadratic rates of branching
% and coalescing. Later, we will show how the graph structure and breakpoints
% associated with
% recombination nodes provides sufficient information to later recover the marginal
% trees and calculate sampling probabilities. However, this process traces out
% a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
% and is consequently considerably less efficient \citep{wiuf1999ancestry}.

\section*{The Event ARG data structure}\label{eARG}
Initial work on \emph{simulating} ancestral graphs with recombination was undertaken by
\citet{hudson1983properties,hudson1990gene}, who extended the classical coalescent
framework~\citep{kingman1982coalescent,kingman1982genealogy, hudson1983testing, tajima1983evolutionary}
to include recombination. Hudson's algorithm has led to powerful simulation
tools~\citep{hudson2002generating,kelleher2016efficient, baumdicker2021efficient}, and is
also relevant to efficient analysis, as we shall see later in the
``\nameref{Ancestry_resolution}'' section.

% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?

The precise term ``Ancestral Recombination Graph'' or ARG, was subsequently introduced by
Griffiths and colleagues~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Rather than simulation, their work focused on mathematical results about the underlying branching-coalescing
stochastic process. However, some confusion has arisen from their use of the word ``ARG'' to refer to
two distinct concepts: the process itself, and the graph structure it generates.

%% NB: I (yan) have commented out the following section as the "latent parameter" quote is
%% used again in the "ARG reconstruction" section, and may be more detail than is required here
%% This also loses the description of ARG space being large etc.
%Early work on ARG inference focused on the problem of
%inferring parameters of the
%stochastic process, where the ancestry is regarded as a
%latent parameter to be averaged out
%\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
%fearnhead2001estimating}.
%These methods met with limited success
%because the state space of ARGs is overwhelmingly large and
%lacks a simple geometry or neighbourhood structure for inference or
%sampling methods to  exploit.
%The focus subsequently shifted to the more
%tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
%the minimum number of recombinations required
%to explain an input dataset [CITATIONS], and inferring the corresponding
%ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
%This highlighted the distinction between the ARG as a stochastic process
%and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
%since parsimonious ARGs have no immediate connection to
%the stochastic process.

The second concept, that of the ARG being a \emph{data structure} was adopted by later authors,
when focussing on the problem of inferring an ARG-like structure (or a minimum number of
recombinations [CITATIONS]) without assuming a specific stochastic
model, for example using parsimony~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
Following~\citet{minichiello2006mapping}, we adopt this convention, using ARG
to mean an ancestral graph structure, without reference to the process that generated it.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Immediate edge annotation, reflecting ``local'' transmission of all
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotation: the Hudson-like simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
Line widths reflect the genomic span of each edge, which
helps visualise the relative importance of different transmission pathways.
}
\end{figure}

@@

% NB: below is a discussion of the Hudson algorithm which we might want to
% retain in an appendix
% Each lineage consists of a list of
% disjoint ancestry segments $(\ell, r, a)$, where
% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
% tracking the number of samples to which the lineage is ancestral over that interval.
% (We also usually track the tree node associated with each segment, but
% that is not important for our purposes here so we omit it.)
% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
% the present day as a series of random common ancestor or recombination events.
% Recombination events occur at a rate determined by the amount of ancestral material and
% the way in which it is distributed along a chromosome.
% Let $L$ be the set of ancestors at a given time $t$. Recombination events
% happen at rate $\rho \nu / (m - 1)$ where
% \[
% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
% \]
% is the number of available `links' that may be broken. Thus, the rate of
% recombination is determined by the left- and right-most extent of the
% ancestral material carried by each lineage. At a recombination
% event we choose one of these links uniformly and break it. Given a lineage
% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
% $x_1$ and $x_2$ such that FILL IN DETAILS

@@

% The state-space of the Griffiths process is much simpler than Hudson's algorithm,
% which greatly facilitates mathematical reasoning. This simplicity comes at a
% substantial cost, however, if we wish to use it as a practical means of
% simulating recombinant ancestry. The big ARG is vast, and any realisation
% for even moderate levels of recombination is far too large to be of practical
% use. The number of events in the big ARG all the way back to the GMRCA
% is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
% of events required to simulate the little ARG is
% $O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
% This disparity in the number of events in the two formulations is
% because the majority of the events that occur in the big ARG do
% not affect the genetic ancestry of the sample in any way. Recombination
% events that occur outside of ancestral material do not have any bearing
% on the ancestry of the sample, and so the structure is hugely redundant.
% As~\cite{wiuf1999recombination} note,
% ``an `ancestral' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.''
% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

@@

@@

>>>>

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line}
% Do we mean "despite the fact that analysis of genetic data relies on methods
% which ultimately derive from the inheritance paths defined by genetic trees
% and the mutations upon them, analysis so far has focussed on summaries ...
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

@@

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\end{enumerate}

@@

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph
(with particular properties; see section XXX for details),
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to describe recombinant ancestry as single object that
could be inferred in a way
that was missing from Hudson's treatments.

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now broken,
since such parsimonious ARGs are extremely unlikely under the
% JK: I assume this is true; any citations to back it up I wonder?
coalescent with recombination.
\cite{minichiello2006mapping} made this separation of the ARG as a
stochastic process vs the ARG as a data structure explicit, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as describing algorithms for more general phylogenetic
networks~\citep{huson2010phylogenetic}.

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
However, in parallel with this tendancy to decouple the term
Ancestral Recombination graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model an ensemble of correlated
potential ancestral trees and has been of considerable
and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
A vital distinction between the ARG (process) and ASG, however,
is that there is a hard distinction between the stochastic process
(the graph) and an observable realisation (a single tree) in the ASG
because the it encodes \emph{potential} ancestral trees.
Similarly, the Ancestral Gene Transfer Graph
models horizontal gene transfer~\citep{baumdicker2014infinitely}
by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely derived from the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events.
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

\section*{The Event ARG data structure}\label{eARG}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig.~1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Immediate edge annotation, reflecting ``local'' transmission of all
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotation: the Hudson-like simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
Line widths reflect the genomic span of each edge, which
helps visualise the relative importance of different transmission pathways.
}
\end{figure}

@@

@@

@@

\item Throughout, we considered ARGs defined for a single chromosome comprising
of a fixed number of discrete sites. Our argument applies equally well to ARGs
defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
koskela2019robust} and to idealised, continuous chromosomes.

\setcounter{secnumdepth}{2} % Print out appendix section numbers

\appendix
\section*{Appendix}

\subsection*{The Big and Little ARG}
\label{app-big-and-little-arg}

@@

[JK: this is text roughly related to big vs little ARG taken from various
iterations]

In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

% % This is a discussion of Hudson's alg left over from earlier version. We
% % might wish to adapt some of this when defining the Little ARG.

% Each lineage consists of a list of
% disjoint ancestry segments $(\ell, r, a)$, where
% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
% tracking the number of samples to which the lineage is ancestral over that interval.
% (We also usually track the tree node associated with each segment, but
% that is not important for our purposes here so we omit it.)
% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
% the present day as a series of random common ancestor or recombination events.
% Recombination events occur at a rate determined by the amount of ancestral material and
% the way in which it is distributed along a chromosome.
% Let $L$ be the set of ancestors at a given time $t$. Recombination events
% happen at rate $\rho \nu / (m - 1)$ where
% \[
% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
% \]
% is the number of available `links' that may be broken. Thus, the rate of
% recombination is determined by the left- and right-most extent of the
% ancestral material carried by each lineage. At a recombination
% event we choose one of these links uniformly and break it. Given a lineage
% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
% $x_1$ and $x_2$ such that FILL IN DETAILS
